Pattern matching for\\ $(231,213)$-avoiding permutations

### 2015-01-31 article.tex
introduction

\DeclareMathOperator{\RUN}{run}

@@

\title{Pattern matching for $(231,213)$-avoiding permutations}
\author{Both Emerite NEOU, Romeo RIZZI}
\date{}

@@

	There exist many generalisation of patterns that are worth considering
	in the context of algorithmic issues in pattern involvement
	(see \cite{Kitaev:book:2011} for an up-to-date survey).
	Vincular patterns, also called generalized patterns,
	resemble (classical) patterns, with the constraint that some of the letters in
	an occurrence must be consecutive.
	Of particular importance in our context,
	Bruner and Lackner \cite{Bruner:Lackner:SWAT:2012}
	proved that deciding whether a vincular pattern
	$\sigma$ of length $k$ occurs in a permutation
	$\pi \in \mathfrak{S}_n$ is $W[1]$-complete for
	parameter $k$.
	bivincular patterns generalize classical patterns even further than vincular
	patterns. Indeed, in bivincular patterns, not only positions but also values
	of elements involved in a matching may be forced to be adjacent

@@

	\section{Definition}

		This section presents the decomposition used and some notation.\\

		We denote by $\Av{n}{213,231}$ the class
		of permutation of size $n$ which avoid 
		the motif $213$ and $231$. 
		$\Perm{n}$ the permutation of size $n$.
		$[n]$ the set $\{1,2,...,n\}.$
		For a permutation p, let $p[i:j]$, 
		be the permutation from element $i$ to $j$,
		$p[:j]$ be the prefix ending at index $j$
		and $p[i:]$ be the suffix starting at index $i$.\\

		Given two permutations
		$\pmotif$ and $\ptext$,
		we say that $\pmotif$
		is a motif/match of
		$\ptext$,
		if and only if
		there exists 
		and increasing function 
		$\sigma:[|\pmotif|] \rightarrow [|\pmotif|]$
		such that : $\forall i$ and 
		$\forall j$ $i,j \in |\pmotif| $
		if $\pmotif(i)<\pmotif(j)$
		then $\ptext(\sigma(i))<\ptext(\sigma(j))$.\\

		%\subsection{Stripe}

>>>>

\DeclareMathOperator{\RUN}{run}
\DeclareMathOperator{\RED}{red}
\DeclareMathOperator{\AV}{Av}

@@

\title{Pattern matching for\\ $(231,213)$-avoiding permutations}
\author{Both Emerite NEOU, Romeo RIZZI}
\date{}

@@

	There exist many generalisation of patterns that are worth considering
	in the context of algorithmic issues in pattern involvement
	(see \cite{Kitaev:book:2011} for an up-to-date survey).
	\emph{Vincular patterns}, also called 
	\emph{generalized patterns},
	resemble (classical) patterns, with the constraint that some of the letters in
	an occurrence must be consecutive.
	Of particular importance in our context,
	Bruner and Lackner \cite{Bruner:Lackner:SWAT:2012}
	proved that deciding whether a vincular pattern
	$\sigma$ of length $k$ occurs in a permutation
	$\pi \in \mathfrak{S}_n$ is $W[1]$-complete for
	parameter $k$.
	\emph{Bivincular patterns} generalize classical patterns even further 
	than vincular
	patterns. Indeed, in bivincular patterns,
	not only positions but also values
	of elements involved in a matching may be forced to be adjacent

	This paper is organized as follows.

@@

\section{Definition}
\label{section:Definitions}

A \emph{permutation} of length $n$ is a one-to-one function from an
$n$-element set to itself.
We write permutations as words 
$\sigma = \sigma_1\,\sigma_2\,\ldots\,\sigma_n$, whose letters are distinct
and uasually consist of the integers $1\,2\,\ldots\,n$.
For the sake of convenience, 
we let 
$\pi[i:j]$ stand for 
$\pi_i\,\pi_{i+1}\,\ldots\,\pi_j$,
$\pi[:j]$ stand for $\pi[1:j]$ and
$\pi[i:]$ stand for $\pi[i:n]$.
We let $S_n$ denote the set of all permutations of length $n$.

% The \emph{reduced form} of a permutation $\sigma$ on a set
% $\{j_1, j_2, \ldots, j_k\}$ where
% $j_1 < j_2 < \ldots < j_n$ is the permutation $\sigma'$
% obtained by renaming the letters of $\sigma$ so that 
% $j_i$ is renamed $i$ for all $1 \leq i \leq k$.
% We let $\RED(\sigma)$ denote the reduced form of $\sigma$.
% For example $\RED(453) = 231$
% while $\RED(3174) = 2143$.

A permutation $\pi$ is said to \emph{contain} the permutation $\sigma$ 
if there exists a subsequence of (not necessarily consecutive) 
entries of $\pi$ that has the same relative order as $\sigma$, 
and in this case $\sigma$ is said to be a \emph{pattern} of 
$\pi$, written $\sigma \preceq \pi$. 
Otherwise, $\pi$ is said to \emph{avoid} the permutation $\sigma$. 
For example, the permutation $\pi = 3\,9\,1\,8\,6\,7\,4\,5\,2$ 
contains the pattern $\sigma = 51342$, 
as can be seen in the highlighted subsequence of 
$\pi = 3\,\mathbf{9}\,\mathbf{1}\,8\,\mathbf{6}\,\mathbf{7}\,\mathbf{4}\,5\,2$
(or 
$\pi = 3\,\mathbf{9}\,\mathbf{1}\,8\,\mathbf{6}\,\mathbf{7}\,4\,\mathbf{5}\,2$
or
$\pi = 3\,\mathbf{9}\,\mathbf{1}\,8\,\mathbf{6}\,\mathbf{7}\,4\,5\,\mathbf{2}$
). 
Each subsequence $(9\,1\,6\,7\,4$, 
$9\,1\,6\,7\,5$, 
$9\,1\,6\,7\,2$) is called a 
\emph{copy}, 
\emph{instance}, or 
\emph{occurrence} of $\sigma$. 
Since the permutation $\pi = 3\,9\,1\,8\,6\,7\,4\,5\,2$  contains no increasing subsequence of length four, $\pi$ avoids $1\,2\,3\,4$.

Suppoe $P$ is a set of permutations. We let $\AV_n(p)$ denote the
set of all $n$-permutations avoiding each permutation in $P$.
For the sake of convenience, we omit $P$'s braces thus having
e.g. $\AV_3(132,4312)$ instead of
$\AV_3(\{132,4312\})$.
A basic example is if 
$\pi = \pi_1\,\pi_2\,\ldots\,\pi_n$ avoids $3\,2\,1$, i.e., 
has no decreasing subsequence of length $3$, then its reverse, 
$\pi' = \pi_n\,\pi_{n-1}\,\ldots\,\pi_1$ avoids $1\,2\,3$, i.e., 
has no increasing subsequence of length $3$.

A \emph{bivincular pattern} (abbreviated BVP) $\sigma$
of length $k$ is a permutation in $S_k$ written in
two-line notation 
(that is the top row is $1\,2\,\ldots\,k$ and the bottom row
is a permutation $\sigma_1\,\sigma_2\,\ldots\,\sigma_k$).
We have the following conditions on the top and bottom rows
of $\sigma$:
\begin{itemize}
	\item
	If the bottom line of $\sigma$ contains
	$\underline{\sigma_i\,\sigma_{i+1}\,\ldots\,\sigma_j}$
	then the letters corresponding to 
	$\sigma_i\,\sigma_{i+1}\,\ldots\,\sigma_j$ in an occurrence of
	$\sigma$ in a permutaton must be adjacent, whereas there is
	no adjacency condition for
	non-underlined consecutive letters.
	Moreover if the bottom row of $\sigma$ begins with
	$_\llcorner{\sigma_1}$ then any occurrence of $\sigma$ 
	in a permutation $\pi$ must begin with the leftmost 
	letter of $\pi$,
	and 
	if the bottom row of $\sigma$ begins with
	${\sigma_k}_\lrcorner$ then any occurrence of $\sigma$ 
	in a permutation $\pi$ must end with the rightmost 
	letter of $\pi$.
	\item
	If the top line of $\sigma$ contains
	$\overline{i\,i+1\,\ldots\,j}$ then the letters corresponding to
	$\sigma_i, \sigma_{i+1}, \ldots, \sigma_j$ in an
	occurrence of $\sigma$ in a permutation must be adjacent in values,
	whereas there is no value adjacency restriction for non-overlined 
	letters.
	Moreover, if the top row of $\sigma$ begins with
	$^ulcorner{1}$ then
	any occurrence of $\sigma$ is a permutation $\pi$ must begin with
	the smallest letter of $\pi$, and
	if top row of $\sigma$ ends with $k^\urcorner$ then
	any occurrence of $\sigma$ is a permutation $\pi$ must end with
	the largest letter of $\pi$.
\end{itemize}
